---
title: "Aggregate Site table from PSM table"
author: "Moritz Madern"
date: "2022-08-24"
output: html_document
---

Data input:
1) modified PSM-table (output of "IM.Rmd" script)
2) proteins table (potentially filtered already)

Data output:
1) MaxQuant proteingroups.txt


This script performs aggregation of PSMs to higher-level site features by a) summation of reporter intensities, b) calculating weighted averages to obtain aggregated EIL values, c) and more (e.g. filtering, aggregating PPF, visualization). 


```{r Load required packages etc.}

## Load packages
library(tidyverse)


## Create Results folder
if (!file.exists("Results")){
  dir.create("Results")
}

```


```{r Specify required parameters}

## Save working directory
wd <- getwd()


## 1) Parameters specific to the PSM table:

## Specify filepath to PSM table containing: a) reporter ion intensity columns (impurity-corrected and normalized); b) columns providing feature-wise EIL-values (estimated interference level) and PPF-values (precursor purity fraction); a unique id by which other tables (i.e. Site table) refer to it.
filepath_PSM_table = "Results/PSM.txt"

## Specify reporter ion pattern in PSM table (expected to be impurity corrected and normalized)
reporter_pattern_PSM = "reporters_[0-9]*(|N|C)_norm$"

## Specify EIL column name in PSM table
eil_columnname = "EIL"

## Specify PPF column name in PSM table
ppf_columnname = "PPF"

## Specify minimum PPF filter threshold
ppf_threshold = 0

## Specify the name of the id column by which other tables refer to the PSM table
id_columnname_PSM_table = "id"

## Specify the name of the column containing the integer number of modifications for the respective PSM precursor peptide
number_modifications_columname = "Acetyl..K."  

## Specify the name of the column containing the minimum recorded intensity in a PSM's MS2 scan
min_ms2_int_columnname = "minIntensity_MS2"

## Specify if remaining NAs in PSMs should be substituted with minimum MS2 intensity
substitute_NAs_with_min_intensity = FALSE




## 2) Parameters specific to the site table and subsequent filtering of sites:

## Specify filepath to  site table (e.g. acetyl-site table , phospho-site table) that lists features (as rows) of a higher aggregation level than PSMs. In this script, PSMs in the provided PSM table will be aggregated (by summation of PSM reporter ion intensities) to those features. Naturally, this requires the information of which PSMs are allotted to which feature - in MaxQuant, this information is provided by columns like "MSMS IDs". In parallel, the EIL-values will also be aggregated (as the weighted average of the PSMs, where weights are proportional to total PSM reporter ion intensity). On that basis, we can later perform site-to-protein normalization (by another script).
filepath_feature_table =  "./proteinGroups.txt"

## Specify MSMS id column name in the feature table. This column points to the every individual PSM that count to a single feature (separated by ";"). Based on this information, aggregation is performed.
msms_id_columnname_feature_table = "MS.MS.IDs" 

## Specify minimum top 3 intensity filter threshold on site-level
top3_threshold_feature_table = 0


```


```{r Read in PSM table and derive relevant objects}

## Read in PSM table
df_PSM <- read.delim(filepath_PSM_table, header=TRUE, sep="\t")


## Print the head of PSM reporter intensities (to check if specified pattern is correct!)
df_PSM[,grepl(names(df_PSM), pattern=reporter_pattern_PSM)] %>% head()


## Rename EIL column of PSM table to "EIL"
names(df_PSM)[names(df_PSM) == eil_columnname] <- "EIL"


## Rename PPF column of PSM table to "PPF"
names(df_PSM)[names(df_PSM) == ppf_columnname] <- "PPF"


## Rename id column name of PSM table to "id"
names(df_PSM)[names(df_PSM) == id_columnname_PSM_table] <- "id"


## Perform additional filtering of PSMs based on PPF
writeLines(paste0("PSMs before filtering based on PPF: ", nrow(df_PSM)))
df_PSM <- df_PSM[df_PSM$PPF >= ppf_threshold,]
writeLines(paste0("\nPSMs after filtering based on PPF: ", nrow(df_PSM)))


## print the PPF threshold used for filtering
writeLines(paste0("\nThe chosen PPF threshold used for filtering: ", ppf_threshold))

```

```{r Read in protein table and derive relevant objects. Also Filter out reverse and CONs, + based on score. Extract gene names from fasta if empty}

## Read in feature table
df_feature <- read.delim(filepath_feature_table, header=TRUE, sep="\t")


## Filter feature table (assuming it is a MaxQuant output): remove CONs and reverse
writeLines("Filtering rows based on various criteria. Number of rows:")
writeLines(paste0("Before filtering anything: ", nrow(df_feature)))
df_feature <- df_feature[df_feature$Potential.contaminant == "",]
df_feature <- df_feature[df_feature$Reverse=="",]
writeLines(paste0("After filtering contaminants and reverse hits: ", nrow(df_feature)))


## Filter for at least 2 razor + unique peptides
bool_keep <- df_feature$Razor...unique.peptides >= 2
df_feature <- df_feature[bool_keep,]
writeLines(paste0("After filtering for at least 2 razor+unique peptides: ", nrow(df_feature)))


## Filter out only identified by site
bool_keep <- df_feature$Only.identified.by.site != "+"
df_feature <- df_feature[bool_keep,]
writeLines(paste0("After filtering out proteins only identified by site: ", nrow(df_feature)))




## Check gene name column
writeLines("\nExtracting Gene names from fasta header:")
writeLines(paste0("Number of Gene name entries that are empty: ", sum(df_feature$Gene.names == "")))


## Extract Fasta Headers, and within them, gene names.
gene_names <- character(nrow(df_feature))        #initialize final vector
fasta_headers <- df_feature$Fasta.headers
fasta_headers <- ifelse(substring(fasta_headers, first=1, last=1)==";",
                          no=fasta_headers, 
                          yes=substring(fasta_headers, first=regexpr(fasta_headers,pattern="[A-Za-z0-9]"), last=10000))
fasta_headers_split <- strsplit(fasta_headers, split=";")
for (i in 1:length(fasta_headers_split)){
  # in case protein i had a valid entry already, take the original entry
  if ("Gene.names" %in% names(df_feature) && !df_feature$`Gene.names`[i]== ""){
    gene_names[i] <- df_feature$Gene.names[i]
    next()
  }
  fasta_headers_split_i <- fasta_headers_split[[i]]
  if (length(fasta_headers_split_i) > 0 && grepl(fasta_headers_split_i,pattern="GN=")){
      gene_names_split_i <- substring(fasta_headers_split_i ,first= regexpr(fasta_headers_split_i,pattern="GN=") + 3, last=nchar(fasta_headers_split_i))
  } else{
      gene_names_split_i <- substring(fasta_headers_split_i ,first = 1, last=nchar(fasta_headers_split_i))
  }
  gene_names_split_i
  gene_names_split_i <- substring(gene_names_split_i, first=1, last=regexpr(gene_names_split_i, pattern=" |$")-1)
  gene_names_split_i
  if(length(gene_names_split_i)>1){
    gene_names_i <- paste0(gene_names_split_i, collapse=";")
  } else{
    gene_names_i <- gene_names_split_i
  }
  gene_names[i] <- ifelse(length(gene_names_i)>0, yes=gene_names_i, no="")
}

## replace old vector with new gene name vector
df_feature$Gene.names <- gene_names
writeLines(paste0("Number of Gene name entries that are empty after matching with fasta header: ", sum(df_feature$Gene.names == "")))

```

```{r Perform aggregation of reporter ion intensities and EIL values from PSM-table to site table based on ID reference columns}

## Extract msms id column in feature table
msms_id_pointer <- df_feature[,msms_id_columnname_feature_table]


## Initiate aggregated reporter intensity matrix
m_agg_reporterInt <- matrix(numeric(1),
                            ncol=sum(grepl(names(df_PSM), pattern=reporter_pattern_PSM)),  # number of reporter ion columns in PSM-table
                            nrow=nrow(df_feature))
colnames(m_agg_reporterInt) <- paste0(grep(names(df_PSM), pattern=reporter_pattern_PSM, value = TRUE), "__aggregated")
 

## Initiate aggregated EIL vector and PPF vector
v_agg_eil <- numeric(nrow(df_feature))
v_agg_ppf <- numeric(nrow(df_feature))


## Initiate vector indicating whether PSMs for the respective site could be found in the PSM table
v_agg_control <- rep("+", times=nrow(df_feature))


## Go over each site i in df_feature and aggregate their reporter intensities + EIL values  + PPF values from the PSM-table
for (i in 1:nrow(df_feature)){
  
  # extract corresponding PSM ids
  psm_ids_i <- msms_id_pointer[i]
  
  # split them at ";"
  psm_ids_i <- strsplit(psm_ids_i, split=";")[[1]]
  
  # check if the feature really points to PSMs in the PSM-table
  df_PSM_i <- df_PSM[df_PSM$id %in% psm_ids_i,]
  if (nrow(df_PSM_i)==0){ 
    v_agg_control[i] <- "-"
    next()
  }
  
  # extract reporter intensity columns
  m_PSM_i <- df_PSM_i[,grepl(names(df_PSM_i), pattern=reporter_pattern_PSM)] %>% as.matrix(.)
  
  # substitute 0s with NAs
  m_PSM_i[m_PSM_i==0] <- NA
  
  # calculate rowSums to infer weights in EIL aggregation + PPF aggregation
  rowsums_i <- rowSums(m_PSM_i, na.rm=TRUE)
  weights_i <- rowsums_i/sum(rowsums_i)
  
  # aggregate EIL and PPF
  v_agg_eil[i] <- sum(weights_i * df_PSM_i$EIL)
  v_agg_ppf[i] <- sum(weights_i * df_PSM_i$PPF)
  
  # substitute NAs and zeros with minimum intensity in MS2 scan if specified
  if (substitute_NAs_with_min_intensity){
    for (j in 1:nrow(m_PSM_i)){
      row_j <- m_PSM_i[j,]
      min_j <- df_PSM_i[j,min_ms2_int_columnname]
      row_j[is.na(row_j) | row_j == 0] <- min_j
      m_PSM_i[j,] <- row_j
    }
  }
  
  # calculate column sums (i.e. aggregation from PSMs to feature by summation)
  m_agg_reporterInt[i,] <- colSums(m_PSM_i, na.rm = TRUE)
}


## merge the results to df_PSM
df_feature <- cbind(df_feature, m_agg_reporterInt)
df_feature$EIL <- v_agg_eil
df_feature$PPF <- v_agg_ppf
df_feature$agg_control <- v_agg_control

```


```{r Filter out features coming from previously filtered PSMs }

## Note: This step filters out sites that wholly derive from PSMs which have previously been filtered out in the PSM table (due to, for example: 0 isobaric labels (i.e. unlabeled); second peptides; insufficient PPF, insufficient valid values per Group)


## get an overview of how much is filtered out
barplot(table(df_feature$agg_control), border="grey", col="grey", ylab = "frequency", xlab="passed filter", main="Overview of filtering")


## perform filtering
writeLines(paste0("Before integrating PSM-filters: ", nrow(df_feature)))
df_feature <- df_feature[df_feature$agg_control == "+",]
writeLines(paste0("After integrating PSM-filters: ", nrow(df_feature)))

```

```{r Filter out features below the 1% percentile average top3 intensities}

# ## Extract aggregated intensities
# m_aggregated <- df_feature[,grepl(names(df_feature), pattern="aggregated")] %>% as.matrix()
# 
# 
# ## calculate average top 3 intensity per feature
# mean_top3 <- apply(m_aggregated, MARGIN = 1, FUN=function(x){
#                                   x_sort <- sort(x, decreasing = TRUE)
#                                   if(length(x_sort)>=3){
#                                     res <- mean(x_sort[1:3])
#                                   } else{
#                                     res <- mean(x_sort[1:length(x_sort)])
#                                   }
#                                   return(res)
#                                   })
# 
# 
# ## filter out sites below average top3 threshold
# hist(log2(mean_top3), breaks=100, border="grey", col="grey", main="distribution of average top3 log2 intensities")
# percentile_1 <- quantile(mean_top3, probs=0.01, na.rm=TRUE)
# abline(v=log2(percentile_1), col="red", lty="dashed", lwd=2)
# bool_keep <- mean_top3 >= percentile_1
# writeLines(paste0("Number of sites after applying 1% percentile average top3 filter threshold: ", nrow(df_feature)))
# df_feature <- df_feature[bool_keep,]
# writeLines(paste0("\nNumber of sites after applying 1% percentile average top3 filter threshold: ", nrow(df_feature)))
# writeLines(paste0("Chosen average top3 intensity threshold for filtering: ", percentile_1))

```


```{r Export results}

## If no specifc output name was given, name resulting table based on input file's name
ind_substring <- rev(unlist(gregexpr(filepath_feature_table, pattern="[\\]|[/]")))[1]
outputname_path <- paste0("Results/aggregated_",substring(filepath_feature_table, first = ind_substring + 1, last=nchar(filepath_feature_table))) 


## Generate output table
write.table(df_feature, file = outputname_path, sep = "\t", col.names = TRUE, row.names=FALSE, quote=FALSE)


## Report about the export
writeLines("The result table can be found in:")
print(outputname_path)

```


```{r}

```



